Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-23 Manuscript/manuscript.Rmd
manuscrpt edits

<!--Additional threshold-free metrics where also reported including the AUROC and AUPRC.-->

<!--
Lastly, for the task of identifying the genomic annotations that were most predictive of the formation of TAD boundaries, we ranked the variable importances associated with each feature of the ensemble system-specific random forest model. To calculate variable importances for each model, the F1-Score on the out-of-bag portion of the data within each tree was recorded. Then the same was done after permuting each predictor variable. The difference between the two AUPRC values was then averaged over all trees, and normalized by the standard error of the differences. Annotations with low rankings correspond to functional epigenomic features that are either positively or negatively enriched near TAD boundaries, and thus, better allow the model to discriminate between genomic bins that overlap with flanked TAD boundaries.
-->

## ENSEMBLE FRAMEWORK

An ensemble framework was then established for the purposes of systematically creating a variety of models that each included different sets of parameters described above <!--(Figure 2)-->. Each system in the framework is assumed to be composed of a model constructed from a particular cell line and data resolution, with a particular predictor type for the feature space. The data sets used for modeling were composed of the full set of genomic bins that either overlapped with a flanked TAD boundary or did not. The data was then split into a 7:3 ratio of training to testing sets. For the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison. Furthermore, this process was repeated, and performances were aggregated across each autosomal chromosome (omitting chromosome 9) using means and standard deviations.

@@

## THE PROPOSED MODEL

We illustrate a unified approach toward preparing TAD boundary data for predictive modelling purposes from different cell lines and data resolutions (Figure 1). To enable us to build a model that suitably discriminates between regions on the linear genome that are near TAD boundaries and regions that are not, we created a cell line- and resolution-specific response variable that indicated whether a genomic bin of unit width overlapped with a flanked TAD boundary ($Y=1$), or not ($Y=0$). The unit width of each bin on the linear genome corresponded to the resolution of the contact matrix on which domains were called. Moreover, to account for uncertainty in boundary location, TAD boundaries were flanked on either side by the same unit length as the genomic bins.

@@

## ENSEMBLE COMPARISONS

We then developed an ensemble framework to systematically compare and establish the best set of data level characteristics for predicting TAD boundaries using random forest classifiers [??? Do we need RF?] built on cell line specific functional genomic annotations (Figure 3). TAD boundaries were called using the Arrowhead algorithm [@durand2016juicebox] from replicate Hi-C data for 2 cell lines $L = \{GM12878, K562\}$, at four different resolutions $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. For each classifier, we considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, four re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome, and thus, was repeated repeated 21 times (number of autosomal chromosomes, excluding chromosome 9 [??? In Methods, not here, you should explain why]), $C = \{1, 2, ..., 22\}$. A system produced from the ensemble framework was defined as: $S = \{c, r, p, b\}$, where $l \in L$, $r \in R$, $p \in P$, $b \in B$, and $c \in C$. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods). We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials. [??? This is Methods. What is so compelling about those 'results' for the message you try to prove?] [??? Currently, Methods and the results are redundant. Look how others present their models in the results section, and how methods section is structured in such cases. Specifically, some key definitions, like OC, OP, Distance should be solidly introduced in the results section, while log2 status of distance and supplementary figure 1 description are for methods]

[???probably better to report medians here]
Expectedly, the average number of genomic bins, TADs called, and TAD boundaries increased with the increased resolution of cell-line specific Hi-C data (Table 1; Supplementary Tables 3). The correlation of each cell-line specific quantity with resolution was found to be significant, as evidenced by large intraclass correllation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem was found to be much more consistant across resolutions in both cell lines. Likewise, this relationship was not significantly correlated with resoltuion for either cell line, indicating that there exists a severe disparity between classes at all resolutions of Hi-C data.

[??? Very bad and boring. You should tell the STORY, not the report. E.g., "Expectedly, the average number of genomic bins, TADs, and TAD boundaries increased with the increased resolution of Hi-C data (Table 1). However, the imbalance ratio, defined as ..., remained severe (~XXX) at all resolutions (Table 1). In summary, the problem of class imbalance exists at all resolutions of Hi-C data."; !!! editted]

>>>>

## ENSEMBLE FRAMEWORK & COMPARISONS

To systematically summarize and compare our results, we created an ensemble framework. Each system in the framework was assumed to be composed of a model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. The cell line- and resolution-specific data was then split into a 7:3 ratio of training to testing sets. Furthermore, for the task of performance evaluation and prediction, the test set remained constant for all ensemble systems to account for an unbiased comparison. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome, and thus, was repeated repeated 21 times (number of autosomal chromosomes, excluding chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods). We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

@@

We analyzed publically available Hi-C data sets for the GM12878 and K562 cell lines provided by Rao et al. [@rao20143d]. Topologically associating domains were then identified using the Arrowhead algorithm [@durand2016juicebox] obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolutions, for each cell line respectively. We then established a unified approach toward preparing cell line specific TAD boundary data for predictive modelling purposes, irrespective of data resolutions (Figure 1).

To enable us to build a model that suitably discriminated between regions on the linear genome that were near TAD boundaries and regions that were not, we created a  resolution-specific response variable that indicated whether a genomic bin of unit width overlapped with a flanked TAD boundary ($Y=1$), or not ($Y=0$). The unit width of each bin on the linear genome corresponded to the resolution of the cell line specific contact matrix on which domains were called. Moreover, TAD boundaries were flanked on either side by the same unit length as the genomic bins to account for uncertainty in boundary location.

Expectedly, the average number of genomic bins, TADs called, and TAD boundaries increased with the increased resolution of cell-line specific Hi-C data (Table 1; Supplementary Tables 3). This correlation with resolution was found to be significant, as evidenced by large intraclass correllation coefficients (ICC) (Table 2; Supplementary Tables 4). However, the class imbalance problem was found to be much more consistant across resolutions in both cell lines. The sparsity of TAD boundaries throughout the genome was found to contribute to heavily imbalanced classes. This imbalance was evidenced by low imbalance ratios, defined as the ratio of the number of genomic bins that overlapped with flanked TAD boundaries ($Y=1$), to the much larger number of genomic bins that did not ($Y=0$). This relationship was not significantly correlated with resoltuion for either cell line, indicating that there existed a severe disparity between classes at all resolutions of Hi-C data.
[??? Very bad and boring. You should tell the STORY, not the report. E.g., "Expectedly, the average number of genomic bins, TADs, and TAD boundaries increased with the increased resolution of Hi-C data (Table 1). However, the imbalance ratio, defined as ..., remained severe (~XXX) at all resolutions (Table 1). In summary, the problem of class imbalance exists at all resolutions of Hi-C data."; !!! editted]

@@
